Instant Continuation Marks--Just Add $\lambda$

### 2012-10-15 paper.tex

Continuation marks \cite{clements2006portable} are a programming language feature which generalizes stack inpection. Not only do they dramatically simplify correct instrumentation \cite{clements2001modeling}, they have been used to allow inspection-based dynamic security checks in the presence of tail-call optimization \cite{clements2004tail}, to express aspect-oriented programming in higher-order languagues \cite{tucker2003pointcuts}, and to implement a form of dynamic binding in parameters.

@@

Now consider the reduction of the same program transformed. We apply the transformation just-in-time as we reduce to prevent term size explosion and promote clarity and omit the uninteresting reductions.
\begin{schemedisplay}
Ch[(wcm 0 ((lambda (x) (wcm x (ccm))) 1))]
\end{schemedisplay}

@@

After the application of contextual information, we reach
\begin{schemedisplay}
((C[((lambda (x) (wcm x (ccm))) 1)] TRUE)
 (((lambda (mark-value) (lambda (rest-marks) Cp[((CONS mark-value) rest-marks)]))
   ((C[0] FALSE) Cp[NIL])) ((FALSE Ch[(SND NIL)]) Cp[NIL])))
\end{schemedisplay}
the transformation of a \scheme'wcm' body. No evaluation of this term occurs until the contextual information has been evaluated and applied. In the process of evaluation, the \scheme'mark-expr' is evaluated and prepended to the list of continutation marks.

This ultimately results in the following:
\begin{schemedisplay}
((C[((lambda (x) (wcm x (ccm))) 1)] TRUE) Cp[((CONS 0) NIL)])
\end{schemedisplay}

It is evident that this term will behave exactly as a top-level term except as this contextual information influences it, and this is exactly the property we have strived for. Expansion of this term yields
\begin{schemedisplay}
(((lambda (flag)
    (lambda (marks)
      (((((C[(lambda (x) (wcm x (ccm)))] FALSE) marks)
         ((C[1] FALSE) marks))
        flag)
       marks))) TRUE) Cp[((CONS 0) NIL)])
\end{schemedisplay}
the expansion of an application. Both these terms are values, so are unaffected by the contextual information.

The elided reduction has the effect of preparing the terms for application.
\begin{schemedisplay}
((((lambda (x) C[(wcm x (ccm))])
   1) TRUE) Cp[((CONS 0) NIL)])
\end{schemedisplay}
reduces to
\begin{schemedisplay}
((C[(wcm 1 (ccm))]
   TRUE) Cp[((CONS 0) NIL)])
\end{schemedisplay}

@@

With the contextual information given, the evaluation of \scheme'ccm' is simple:
\begin{schemedisplay}
((lambda (marks) marks)
 Cp[((CONS 1) NIL)])
\end{schemedisplay}
reduces to
\begin{schemedisplay}
Cp[((CONS 1) NIL)]
\end{schemedisplay}
and we are left with just what we hoped for.

@@

Because of space limitations, we offer only a sketch of the proof here. [The full proof can be found in a technical report at ...]

@@

Expressing continuation marks in a monadic way is simple enough. In Haskell,
\begin{lstlisting}[language=Haskell]
data CM m a = CM ((Bool,[m]) -> a)

@@

runCM :: CM m a -> a
runCM (CM m) = m (False,[])
\end{lstlisting}
is one way, using our approach. The factorial functions leveraging this implementation exhibit the correct continuation mark behavior but this is \emph{not} a monad. Specifically, the right identity law, \texttt{m >>= return = m}, does not hold since \texttt{CM} conceptually adds a stack frame in bind (\texttt{>>=}). Ager et al. \cite{ager2003functional} implemented a similar stack-inspection monad in terms of a lifted state monad, but resorted to manual management of the stack frames.

\section{Future Work and Conclusion}

@@

Continuation marks support a bevy of instrumentation tools and advanced language features in a generalized, portable way. Despite their demonstrated utility, they have not yet found their way into most languages. A verified characterization of continuation marks in a pure computational language provides implementors of higher-order languages a correct compiler for continuation marks.

>>>>

Continuation marks \cite{clements2006portable} are a programming language feature which generalizes stack inspection. Not only do they dramatically simplify correct instrumentation \cite{clements2001modeling}, they have been used to allow inspection-based dynamic security checks in the presence of tail-call optimization \cite{clements2004tail}, to express aspect-oriented programming in higher-order languages \cite{tucker2003pointcuts}, and to implement a form of dynamic binding in parameters.

@@

Now consider the reduction of the same program transformed. We apply the transformation just-in-time as we reduce to prevent term size explosion and promote clarity and omit uninteresting reductions.
\begin{schemedisplay}
Ch[(wcm 0 ((lambda (x) (wcm x (ccm))) 1))]
\end{schemedisplay}

@@

After the application of contextual information, we reach
\begin{schemedisplay}
((C[((lambda (x) (wcm x (ccm))) 1)] TRUE)
 (((lambda (mark-value) (lambda (rest-marks) Cp[((CONS mark-value) rest-marks)]))
   ((C[0] FALSE) Cp[NIL])) ((FALSE Ch[(SND NIL)]) Cp[NIL])))
\end{schemedisplay}
the transformation of the \scheme'wcm' body. Terms within are arranged so that correct evaluation occurs within the native call-by-value regime. This evaluates \scheme'mark-expr' and and prepends its value to the list of continuation marks before proceeding with evaluation of \scheme'body-expr'. This reduction soon yields the following term:
\begin{schemedisplay}
((C[((lambda (x) (wcm x (ccm))) 1)] TRUE) Cp[((CONS 0) NIL)])
\end{schemedisplay}
It is evident that this term will behave exactly as a top-level term except as this contextual information influences it, and this is exactly the property we have strived for. Expansion of this term yields
\begin{schemedisplay}
(((lambda (flag)
    (lambda (marks)
      (((((C[(lambda (x) (wcm x (ccm)))] FALSE) marks)
         ((C[1] FALSE) marks))
        flag)
       marks))) TRUE) Cp[((CONS 0) NIL)])
\end{schemedisplay}
the expansion of an application. In this example, both the operator and operand are values, so are essentially unaffected by the application of contextual information; this application has the effect of preparing the terms for application:
\begin{schemedisplay}
((((lambda (x) C[(wcm x (ccm))])
   1) TRUE) Cp[((CONS 0) NIL)])
\end{schemedisplay}
reduces to
\begin{schemedisplay}
((C[(wcm 1 (ccm))]
   TRUE) Cp[((CONS 0) NIL)])
\end{schemedisplay}

@@

Once given the contextual information, the evaluation of \scheme'ccm' is simple:
\begin{schemedisplay}
((lambda (marks) marks)
 Cp[((CONS 1) NIL)])
\end{schemedisplay}
reduces to
\begin{schemedisplay}
Cp[((CONS 1) NIL)]
\end{schemedisplay}
and we are left with just what we hoped for.

@@

Because of space limitations, we offer only a sketch of the proof here. The full proof can be found in a technical report\footnote{\texttt{http://students.cs.byu.edu/\textasciitilde{}krgerman/esop2013-technical-report.pdf}}.

@@

Expressing continuation marks in a monadic way is simple enough. In Haskell,
\begin{minipage}{\linewidth}
\begin{lstlisting}[language=Haskell]
data CM m a = CM ((Bool,[m]) -> a)

@@

runCM :: CM m a -> a
runCM (CM m) = m (False,[])
\end{lstlisting}
\end{minipage}
is one way, using our approach. The factorial functions leveraging this implementation exhibit the correct continuation mark behavior but this is \emph{not} a monad. Specifically, the right identity law, \texttt{{\color{gray}m} >>= {\color{RoyalBlue}return} = {\color{gray}m}}, does not hold since \texttt{{\color{gray}CM}} conceptually adds a stack frame in bind (\texttt{>>=}). Ager et al. \cite{ager2003functional} implemented a similar stack-inspection monad in terms of a lifted state monad, but resorted to manual management of the stack frames.

\section{Conclusion and Future Work}

Continuation marks support a bevy of instrumentation tools and advanced language features in a generalized, portable way. Despite their demonstrated utility, they have not yet found their way into most languages. A verified characterization of continuation marks in a pure computational language provides implementors of higher-order languages a correct compiler for continuation marks which we have demonstrated for JavaScript.

Our macro-style approach would be more useful if deployed as a proper macro, but this requires a hygienic macro system which many languages lack. \texttt{sweet.js} is just such a system for JavaScript in early stages and is ideal for our transformation.

@@
